---
title: "Using MCP Servers with `argent`"
format: html
execute:
  echo: true
  eval: false
---

# Introduction

The Model Context Protocol (MCP) enables LLMs to securely interact with external tools and data sources. This vignette demonstrates how to integrate MCP servers with argent to extend your AI agent's capabilities.

MCP (Model Context Protocol) is an open protocol that standardizes how applications provide context to LLMs. MCP servers expose three main primitives:

- **Tools**: Executable functions the LLM can call
- **Resources**: File-like structured data the LLM can access (not supported yet)
- **Prompts**: Predefined templates for interactions (not supported yet)

There are two main types of MCP servers: *HTTP* and *stdio*. HTTP servers are typically used for cloud-based services like GitHub, while stdio servers are typically used for local services like Docker or npx. Both types of MCP servers can be used with `argent`.

# Setup

```{r}
library(argent)
```

# Basic Usage

## Using an 'HTTP' MCP Server

Let's look at the GitHub MCP server as an example. It is a HTTP server that can be used to interact with the GitHub API.

```{r}
github_mcp_client <- mcp_connect(
    name = "github",
    type = "http",
    url = "https://api.githubcopilot.com/mcp",
    headers = list(Authorization = paste("Bearer", Sys.getenv("GITHUB_PAT")))
)
```

Then, we can get the tools we want from the GitHub MCP server:

```{r}
github_mcp_tools <- mcp_tools(github_mcp_client, tools = c("get_file_contents", "search_code"))
```

Finally, we can call the `get_file_contents` tool manually to see if it works:

```{r}
execute_mcp_tool(
    tool_def = get_mcp_tool(github_mcp_tools, "get_file_contents"),
    arguments = list(owner = "tidyverse", repo = "ellmer", path = "/", ref = "main")
)
```

## Using a 'stdio' MCP Server

Let's look at the BTW MCP server as an example (which requires the [`btw`](https://github.com/posit-dev/btw) package to be installed). 
It is a stdio server that can be used to interact with the `btw` package.

```{r}
btw_mcp_client <- mcp_connect(
    name = "btw",
    type = "stdio",
    command = "Rscript",
    args = c("-e", "btw::btw_mcp_server(tools = btw::btw_tools(c('docs', 'env', 'ide', 'search', 'session')))")
)
```

Then, we can get the tools we want from the `btw` MCP server:

```{r}
btw_mcp_tools <- mcp_tools(
    btw_mcp_client,
    tools = c(
        "btw_tool_session_check_package_installed",
        "btw_tool_docs_available_vignettes",
        "btw_tool_docs_package_help_topics",
        "btw_tool_docs_help_page",
        "btw_tool_docs_vignette",
        "btw_tool_session_package_info",
        "btw_tool_docs_package_help_topics"
    )
)
```

Finally, we can call the `help_topics` tool manually to see if it works:

```{r}
execute_mcp_tool(
    tool_def = get_mcp_tool(btw_mcp_tools, "btw_tool_docs_package_help_topics"),
    arguments = list(package_name = "argent", `_intent` = "Vignettes explaining how to use MCP servers with argent")
)
```

# Argent + MCP Servers

Let's use the GitHub MCP server and the BTW MCP server to ask a complex question about the `mcptools` and `ellmer` packages.

```{r}
google <- Google$new()

google$chat(
    "Has 'posit-dev/mcptools' implemented the ability to use HTTP MCP servers with 'ellmer' ?",
    "Use the `get_file_contents` tool to list the contents of GitHub subdirectories, e.g. with path = '/' or 'dir/'.",
    "Use the `btw` tools to explore the help pages and vignettes of the local installation of the `mcptools` package.",
    model = "gemini-2.5-flash",
    tools = flat_list(github_mcp_tools, btw_mcp_tools)
)
```

```{verbatim}
`posit-dev/mcptools` does not directly implement the ability to use HTTP MCP servers with `ellmer`. When `mcptools` acts as an MCP *client* via `ellmer`, it only supports the local (stdio) protocol. 
To connect to remote (HTTP) MCP servers, the `mcptools` documentation recommends using `mcp-remote`, an external tool (a local stdio MCP server) that converts remote HTTP servers to `mcptools`-compatible local ones. This allows `ellmer` (using the stdio protocol) to interact with remote HTTP MCP servers through `mcp-remote`.
```

```{=html}
<details>
<summary><code>print(google, show_tools = TRUE)</code></summary>
```

```{verbatim}
── [ <Google> turns: 8 | Current context: 3786 | Cumulated tokens: 10034 ] ───────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────


── user [1611 / 2155] ────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────

Has 'posit-dev/mcptools' implemented the ability to use HTTP MCP servers with 'ellmer' ? Use the `get_file_contents` tool to list the contents of GitHub subdirectories, e.g. with path = '/' or 'dir/'. Use the `btw` tools to explore the help pages and vignettes of the local installation of the `mcptools` package.

── System ──

You are a helpful AI assistant. Use your knowledge, the files you have access to, and the tools at your disposal to answer the user's query. You can use your tools multiple times, but use them sparingly. Make parallel tool calls if relevant to the user's query. Answer the user's query as soon as you have the information necessary to answer. Self-reflect and double-check your answer before responding. If you don't know the answer even after using your tools, say 'I don't know'. If you do not have all the information necessary to use a provided tool, use NA for required arguments. Today's date is 2025-11-24

── Tool Definitions ──

• get_file_contents(owner, path, ref, repo, sha): Get the contents of a file or directory from a GitHub repository
• search_code(order, page, perPage, query, sort): Fast and precise code search across ALL GitHub repositories using GitHub's native search engine. Best for finding exact symbols, functions,
  classes, or specific code patterns.
• btw_tool_docs_package_help_topics(package_name, _intent): Get available help topics for an R package.
• btw_tool_docs_help_page(package_name, topic, _intent): Get help page from package.
• btw_tool_docs_available_vignettes(package_name, _intent): List available vignettes for an R package. Vignettes are articles describing key concepts or features of an R package. Returns the
  listing as a JSON array of `vignette` and `title`. To read a vignette, use `btw_tool_docs_vignette(package_name, vignette)`.
• btw_tool_docs_vignette(package_name, vignette, _intent): Get a package vignette in plain text.
• btw_tool_session_check_package_installed(package_name, _intent): Check if a package is installed in the current session.
• btw_tool_session_package_info(packages, dependencies, _intent): Verify that a specific package is installed, or find out which packages are in use in the current session. As a last resort, this
  function can also list all installed packages.

── assistant [2155 / 2155] ───────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────

── Tool Calls ──

• btw_tool_session_check_package_installed(package_name = "mcptools", _intent = "Check if mcptools is installed locally to access its documentation.")

── tool [1757 / 4021] ────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────

• Result from btw_tool_session_check_package_installed:

name: btw_tool_session_check_package_installed
arguments:
  package_name: mcptools
  _intent: Check if mcptools is installed locally to access its documentation.
result: Package `mcptools` version 0.2.0 is installed.


── assistant [1866 / 4021] ───────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────

── Tool Calls ──

• btw_tool_docs_package_help_topics(_intent = "List all help topics for mcptools to find relevant documentation on HTTP MCP servers and ellmer.", package_name = "mcptools")

── tool [2040 / 6248] ────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────

• Result from btw_tool_docs_package_help_topics:

name: btw_tool_docs_package_help_topics
arguments:
  _intent: List all help topics for mcptools to find relevant documentation on HTTP
    MCP servers and ellmer.
  package_name: mcptools
result: |-
  [
    {"topic_id":"client","title":"R as a client: Define ellmer tools from MCP servers","aliases":["client","mcp_tools","mcp_client"]},
    {"topic_id":"mcptools-package","title":"mcptools: Model Context Protocol Servers and Clients","aliases":["mcptools","mcptools-package"]},
    {"topic_id":"server","title":"R as a server: Configure R-based tools with LLM-enabled apps","aliases":["server","mcp_server","mcp_session","server"]}
  ]


── assistant [2227 / 6248] ───────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────

── Tool Calls ──

• btw_tool_docs_help_page(package_name = "mcptools", _intent = "Get the help page for the 'client' topic in mcptools to understand its functionality with ellmer and MCP servers.", topic = "client")

── tool [3359 / 10034] ───────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────

• Result from btw_tool_docs_help_page:

name: btw_tool_docs_help_page
arguments:
  package_name: mcptools
  _intent: Get the help page for the 'client' topic in mcptools to understand its
    functionality with ellmer and MCP servers.
  topic: client
result: |-
  ## `help(package = "mcptools", "client")`

  ### R as a client: Define ellmer tools from MCP servers

  #### Description

  These functions implement R as an MCP *client*, so that ellmer chats can
  register functionality from third-party MCP servers such as those listed
  here: <https://github.com/modelcontextprotocol/servers>.

  `mcp_tools()` fetches tools from MCP servers configured in the mcptools
  server config file and converts them to a list of tools compatible with
  the `⁠$set_tools()⁠` method of ellmer::Chat objects.

  #### Usage

  mcp_tools(config = NULL)

  #### Arguments

  |  |  |
  |----|----|
  | `config` | A single string indicating the path to the mcptools MCP servers configuration file. If one is not supplied, mcptools will look...

── assistant [3786 / 10034] ──────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────

No, `posit-dev/mcptools` itself does not directly implement the ability to use HTTP MCP servers with 'ellmer'. The `mcp_tools()` function within `mcptools`, which allows R to act as an MCP client via 'ellmer', only implements the local (stdio) protocol. However, `mcptools` recommends using `mcp-remote` (an external tool) as an intermediary. `mcp-remote` is a local (stdio) MCP server that can connect to remote (http) servers, converting them to an `mcptools`-compatible local protocol. This allows `mcptools` to indirectly utilize HTTP MCP servers through `mcp-remote`.
```

</details>

# Advanced Usage: making our own MCP servers

## Building a STDIO MCP Server

Let's create our own stdio MCP server for interacting with Zotero's local API. This server will be
usable both from argent and from other MCP clients, like Claude Code.

### Prerequisites

- Zotero 7+ must be running
- Enable "Allow other applications on this computer to communicate with Zotero" in Preferences > Advanced

### Server Implementation

```{=html}
<details>
<summary>Zotero MCP Server Implementation</summary>
```

`r knitr::spin_child(here::here("inst", "examples", "zotero_mcp_server.R"))`

</details>

::: {.callout-note appearance="simple"}

This implementation uses `argent`'s annotation parsing & MCP machinery, which allows us to implement a fully-featured MCP server by simply defining the tools we want to expose as annotated functions.
All the 'boilerplate' code for setting up the MCP server is handled by `argent`.

The Zotero MCP server code is available in the package at `inst/examples/zotero_mcp_server.R`.
:::

### Using the MCP Server

```{r}
zotero_client <- mcp_connect(
    name = "zotero",
    type = "stdio",
    command = "Rscript",
    args = c("-e", "argent::mcp_serve_stdio(system.file('examples/zotero_mcp_server.R', package = 'argent'))")
)

zotero_tools <- mcp_tools(zotero_client)
zotero_resources <- mcp_resources(zotero_client)
zotero_prompts <- mcp_prompts(zotero_client)
```

We could test if the MCP server is working by querying it 'manually':

```{r}
#| code-fold: true
#| code-summary: Testing the Zotero MCP Server manually

# Getting details on all my collections
my_collections <- execute_mcp_tool(get_mcp_tool(zotero_tools, "zotero_get_collections"))

# Getting the key of the Neuroscience collection
neuroscience_collection_key <- jsonlite::fromJSON(my_collections) |>
    dplyr::filter(stringr::str_detect(name, "Neuroscience")) |>
    dplyr::pull(key)

# Getting details on all the items/papers in the Neuroscience collection
neuroscience_items <- execute_mcp_tool(
    get_mcp_tool(zotero_tools, "zotero_get_collection_items"),
    arguments = list(collection_key = neuroscience_collection_key)
)

# Getting a specific paper's key with a keyword from its title
neuroscience_paper_key <- jsonlite::fromJSON(neuroscience_items) |>
    dplyr::filter(stringr::str_detect(title, "cross-modal plasticity")) |>
    dplyr::pull(key)

# Getting the details on the paper
neuroscience_paper_details <- execute_mcp_tool(
    get_mcp_tool(zotero_tools, "zotero_get_item"),
    arguments = list(item_key = neuroscience_paper_key)
)

# Getting the fulltext of the paper (as a string)
neuroscience_paper_fulltext <- execute_mcp_tool(
    get_mcp_tool(zotero_tools, "zotero_get_fulltext"),
    arguments = list(item_key = neuroscience_paper_key)
)

cat(neuroscience_paper_fulltext)
```

Now, let's give the tools to an LLM Agent instead:

```{r}
gemini <- Google$new()

gemini$chat(
    "Can you summarize O'Regan's view on sensory substitution and on how the brain differentiates sensory inputs ?",
    "Find at least 2 papers in my library",
    model = "gemini-2.5-flash",
    tools = zotero_tools
)
```

::: {.callout-tip appearance="simple"}

We can use our new Zotero MCP server with other MCP clients (like Claude Code, Claude Desktop, Gemini CLI, etc):

```{json}
{
  "mcpServers": {
    "zotero": {
      "type": "stdio",
      "command": "Rscript",
      "args": [
        "-e",
        "argent::mcp_serve_stdio(system.file('examples/zotero_mcp_server.R', package = 'argent'))"
      ]
    }
  }
}
```

:::

## Building an HTTP MCP Server

Let's create an HTTP MCP server for searching academic papers using the Semantic Scholar API. 
Unlike stdio servers that communicate via standard input/output, HTTP servers expose a REST endpoint that can be accessed over the network.

::: {.callout-note appearance="simple"}
`argent` provides convenience functions to create HTTP MCP servers, but you can also create your own with `plumber2` or other HTTP server frameworks, independently from `argent`.
:::

### Prerequisites

- A Semantic Scholar API key (not required, but the rate limits are low without one)

### Server Implementation

```{=html}
<details>
<summary>Semantic Scholar MCP Server Implementation</summary>
```

`r knitr::spin_child(here::here("inst", "examples", "semantic_scholar_mcp_server.R"))`

</details>

We can start the server **in a separate R session or terminal** with:

```{r}
#| eval: false

mcp_serve_http(system.file("examples/semantic_scholar_mcp_server.R", package = "argent"), port = 8080)
#> ✔ Starting semantic_scholar_mcp_server MCP server on <http://127.0.0.1:8080>
```

::: {.callout-note appearance="simple"}
**Why start the server separately?** Unlike stdio servers (like Zotero above), which are spawned as child processes by `mcp_connect()` using the provided `command` and `args`, HTTP servers are independent network services. `mcp_connect()` with `type = "http"` simply connects to an already-running server at the specified URL—it cannot start the server for you.
:::

::: {.callout-note appearance="simple"}

This implementation uses `argent`'s annotation parsing & MCP machinery, which allows us to implement a fully-featured MCP server by simply defining the tools we want to expose as annotated functions. All the 'boilerplate' code for setting up the MCP server is handled by `argent`.

If we were to create the same MCP server manually using an HTTP server framework like `plumber2`, here is what it would look like:

```{=html}
<details>
<summary>'Manual' HTTP MCP Server Implementation</summary>
```

`r knitr::spin_child(here::here("inst", "examples", "semantic_scholar_mcp_server_plumber.R"))`

Which we can run **in a separate R session or terminal** with:

```{r}
plumber2::api(system.file("examples/semantic_scholar_mcp_server_plumber.R", package = "argent")) |>
    plumber2::api_run(host = "127.0.0.1", port = 8080, block = TRUE, silent = FALSE, showcase = FALSE)
```

</details>

:::

Then, we can connect to it from our main session:

```{r}
semantic_scholar_client <- mcp_connect(name = "semantic-scholar", type = "http", url = "http://127.0.0.1:8080/mcp")

semantic_scholar_tools <- mcp_tools(semantic_scholar_client)
```

Test the server manually:

```{r}
papers <- execute_mcp_tool(
    get_mcp_tool(semantic_scholar_tools, "search_papers"),
    arguments = list(
        query = "Sensorimotor Contingencies",
        limit = 2L,
        fields = "title,authors,year,abstract,citationCount"
    )
)

# Parse results
papers_df <- jsonlite::fromJSON(papers) |> dplyr::select(paperId, title, year, citationCount)

paper_id <- dplyr::slice_max(papers_df, order_by = citationCount)$paperId

# Getting the details of the first paper
execute_mcp_tool(
    get_mcp_tool(semantic_scholar_tools, "get_paper"),
    arguments = list(paper_id = paper_id, fields = "title,authors,year,abstract")
)
```

Now use it with an LLM agent:

```{r}
gemini <- Google$new()

gemini$chat(
    "Find recent papers (2020+) on Sensory Substitution and Cross-Modal Plasticity and summarize the key innovations",
    "When using the `search_papers` tool, use a limit of 2 by search, to not exceed the rate limit.",
    "If you exceed the rate limit, wait a few seconds before trying again.",
    "Use the `get_paper` tool to get the details of the papers like its abstract.",
    model = "gemini-2.5-flash",
    tools = semantic_scholar_tools,
    output_schema = schema(
        name = "research_papers_summary",
        description = "Information about research papers",
        papers = list(
            type = "[object]*",
            description = "Array of paper objects",
            paper_id = "string* The ID of the paper",
            title = "string* The title of the paper",
            year = "integer* The year of the paper",
            main_findings = "string* The main findings of the paper"
        )
    )
)
```

::: {.callout-tip appearance="simple"}
**HTTP vs stdio servers:**

- **HTTP servers** can be accessed remotely, support multiple concurrent clients, and can be deployed as web services
- **stdio servers** are simpler, work locally only, and are typically one client per server instance
- Both types use the same MCP protocol and tool definitions
:::

::: {.callout-tip appearance="simple"}

We can use our new Semantic Scholar MCP server (either the standalone `plumber2` implementation or the `argent` implementation) with other MCP clients with:

```{json}
{
  "mcpServers": {
    "semantic-scholar": {
      "httpUrl": "http://127.0.0.1:8080/mcp"
    }
  }
}
```
:::
